---
title: "HW05-P1 Adv. Data wrangling & Generalization"
format: 
  html: 
    embed-resources: true
---

## Instructions

**Requirements:**

- Ex 1: Get the desired output
- Ex 1: Used general code (for any dataset input)
- Ex 2: Get the desired output
- Ex 2: Used general code (for any dataset input)


## Overview

You will be recreating core functionality of a few functions in the `janitor` [package](https://cran.r-project.org/package=janitor).



## Exercise 1: `top_levels()`

Look at the [example usage](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#count-factor-levels-in-groups-of-high-medium-and-low-with-top_levels) for the `top_levels()` function, and look at the documentation for this function with `?top_levels`.

Without using `top_levels()`, you should write code to summarize `f`:

- Have the first column of output show the categories that correspond to`Low`, `Medium`, and `High`
- Use functions in the `forcats` package
- Automatically calculate a percentage column
- Do this for the number of levels to include in top and bottom groups to be 2 and 1. 


```{r}
f <- factor(c("strongly agree", "agree", "neutral", "neutral", "disagree", "strongly agree"),
            levels = c("strongly agree", "agree", "neutral", "disagree", "strongly disagree"))
```

Your Goal Output (number of levels to include in top and bottom groups = 2)

```
#>                            f n   percent
#>        strongly agree, agree 3 0.5000000
#>                      neutral 2 0.3333333
#>  disagree, strongly disagree 1 0.1666667
```

Your Goal Output (number of levels to include in top and bottom groups = 1)

```
#>                         f n   percent
#>            strongly agree 2 0.3333333
#>  agree, neutral, disagree 4 0.6666667
#>         strongly disagree 0 0.0000000
#>         
```

You should write code that would work for any data frame, not just the `f` factor.

```{r top_levels}
library(tidyverse)
library(forcats)
f <- factor(c("strongly agree", "agree", "neutral", "neutral", "disagree", "strongly agree"),
            levels = c("strongly agree", "agree", "neutral", "disagree", "strongly disagree"))

n_groups <- 2 
all_levs <- levels(f)
n_total <- length(all_levs)

top_vals    <- all_levs[1:n_groups]
bottom_vals <- all_levs[(n_total - n_groups + 1):n_total]
mid_vals    <- setdiff(all_levs, c(top_vals, bottom_vals))

tibble(f = f) %>%
  mutate(f = fct_collapse(f,
    !!paste(top_vals, collapse = ", ")    := top_vals,
    !!paste(mid_vals, collapse = ", ")    := mid_vals,
    !!paste(bottom_vals, collapse = ", ") := bottom_vals
  )) %>%
  count(f, .drop = FALSE) %>%
  mutate(percent = n / sum(n))
```





## Exercise 2: `get_dupes()`

Look at the [example usage](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#explore-records-with-duplicated-values-for-specific-combinations-of-variables-with-get_dupes) for the `get_dupes()` function, and look at the documentation for this function with `?get_dupes`.

Without using `get_dupes()`,  you should write code that:

- returns the records (and inserts a count of duplicates) of `mtcars` that have duplicated values for `wt` and `cyl`.

```{r}
head(mtcars)
```

Your Goal Output:

```
#>     wt cyl dupe_count  mpg  disp  hp drat  qsec vs am gear carb
#> 1 3.44   6          2 19.2 167.6 123 3.92 18.30  1  0    4    4
#> 2 3.44   6          2 17.8 167.6 123 3.92 18.90  1  0    4    4
#> 3 3.57   8          2 14.3 360.0 245 3.21 15.84  0  0    3    4
#> 4 3.57   8          2 15.0 301.0 335 3.54 14.60  0  1    5    8
```


You should write code that would work for any data frame, not just `mtcars`.


```{r get_dupes}
library(tidyverse)

cols_to_check <- c("wt", "cyl")

mtcars %>%
  group_by(across(all_of(cols_to_check))) %>%
  mutate(dupe_count = n()) %>%
  filter(dupe_count > 1) %>%
  ungroup() %>%
  select(all_of(cols_to_check), dupe_count, everything()) %>%
  arrange(desc(dupe_count), wt)
```


## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- forcats package documentation for merging factor levels.
- dplyr across() documentation for handling dynamic column selection.
- Janitor package vignette on counting factor levels.


**Reflection:**

The forcats documentation was essential for understanding how to collapse factor levels while maintaining the original order, specifically using fct_collapse() to merge groups. I found the use of the bang-bang (!!) and walrus (:=) operators particularly helpful for Exercise 1, as they allowed me to create dynamic labels for the collapsed levels based on the input data. Additionally, using the across() and all_of() functions in the get_dupes exercise helped me write generalized code that can target specific columns without hard-coding names, which is a key skill for building reusable data wrangling pipelines.
