---
title: "HW05-P2 Adv. Data Wrangling + Functions"
format: 
  html: 
    embed-resources: true
---

## Instructions

**Requirements:**

Exercise 1

- Wrote a function that completes all tasks
- At least two test cases together covered whole list (snake_case, % to percent, # to number, remove all non-alphanumeric characters, duplicated var names)
- Wrote desired output before testing function
- Get the desired output (dataset with new names)
- Challenge: duplicated vars changed to var1 and var2 (not using make.unique())

Exercise 2

- Wrote a function that completes all tasks
- At least five test cases together covered whole list (len_x1 < len_x2, len_x1 > len_x2, len_x1 == len_x2, len_larger %% len_smaller == 0, len_larger %% len_smaller != 0
- Get the desired output (2 possible paragraphs text at least)
- Challenge: minimal text: function has only two paragraph text options (where the role of x1 and x2 is figured out by which one is longer) rather than four paragraphs where x1 and x2 are manually typed

Exercise 3

- Wrote a function that completes all tasks
- Test cases together covered whole list (different formats, including from_when, using default from_when, birthday_month_day < from_when_month_day, birthday_month_day > from_when_month_day)
- Get the desired output (no negative time differences)

Exercise 4

- Wrote a function that completes all tasks
- Tested the function with the provided course list (TRUE and FALSE for separate_cross_list)
- Get the desired output (data frame should match the desired output for the test case provided)



## Exercise 1

Look at the [example usage](https://cran.r-project.org/web/packages/janitor/vignettes/janitor.html#clean-dataframe-names-with-clean_names) for the `clean_names()` function, and look at the documentation for this function with `?clean_names`.

Without using `clean_names`, write a general function that will clean the names of a data frame. Your function should:

- **Inputs**
  - Take a dataset as an input and return a dataset as the output

- **Function purpose**:
  - In the output, the variable names should be `snake_case`, meaning `_` instead of spaces in the name.
  - Convert `%` to "percent" and `#` to "number" to retain meaning.
  - Remove non-alphanumeric characters (other than `%` and `#`).
  - Append numbers to duplicated names. (e.g., If there are two variables called `var`, they should be renamed as `var1` and `var2`.)

DON'T worry about:

- Turning `firstName` into `first_name`.
- Handling special characters with accents.

Make sure your function returns the dataset with cleaned names, not just the names themselves. 


```{r clean_names}
library(tidyverse)
clean_names2 <- function(data) {
  nms <- names(data)
  
  nms <- str_replace_all(nms, "%", "percent")
  nms <- str_replace_all(nms, "#", "number")
  
  nms <- str_replace_all(nms, " ", "_")
  
  nms <- str_remove_all(nms, "[^a-zA-Z0-9_]")
  
  new_nms <- character(length(nms))
  for (i in seq_along(nms)) {
    base_name <- nms[i]
    all_instances <- which(nms == base_name)
    if (length(all_instances) > 1) {
      instance_num <- which(all_instances == i)
      new_nms[i] <- paste0(base_name, instance_num)
    } else {
      new_nms[i] <- base_name
    }
  }
  
  names(data) <- new_nms
  return(data)
  }
```

Demonstrate that your function works by showing output on test cases. You must create your test cases and they should cover all the items in the "Your function should:" section. Before the code for testing the functions, write a comment that describes the desired output (the cleaned names) for each test case.

```{r clean_names_test}
test_df_1 <- tibble("% yield" = 1, "# of items" = 2, "simple var" = 3)

test_df_2 <- tibble("var" = 1, "var!!" = 2, "valid_name" = 3)

clean_names2(test_df_1)
clean_names2(test_df_2)
```






## Exercise 2: `recycling_checker()`

- **Inputs:**
    - `x1` and `x2`: two numeric vectors
- **Function purpose:** Print a helpful message to the user warning them if dangerous vector recycling will happen when they try to combine the two vectors with some arithmetic operation.
    - If the vectors are not of the same length, provide informational text. See examples below for how to format this text.
        - Example 1: `x1` is length 4 and `x2` is length 2. Informational text: "x1 and x2 are not the same length. To match the length of x1, x2 will be repeated 2 times. You will not get a warning when you combine these two vectors, but just be wary that vector recycling is occurring."
        - Example 2: `x1` is length 3 and `x2` is length 11. Informational text: "x1 and x2 are not the same length. To match the length of x2, x1 will be repeated 3 times fully with the first 2 elements repeated again. You will get a warning when you combine these two vectors."

```{r recycling_checker}
# Write your function here
recycling_checker <- function(x1, x2) {
  len1 <- length(x1)
  len2 <- length(x2)
  
  if (len1 == len2) {
    return("x1 and x2 are the same length. No recycling will occur.")
  }

  is_x1_longer <- len1 > len2
  L_name <- if(is_x1_longer) "x1" else "x2"
  S_name <- if(is_x1_longer) "x2" else "x1"
  len_L <- max(len1, len2)
  len_S <- min(len1, len2)
  
  times <- len_L %/% len_S
  extra <- len_L %% len_S
  
  if (extra == 0) {
    glue::glue("{L_name} and {S_name} are not the same length. To match the length of {L_name}, {S_name} will be repeated {times} times. You will not get a warning when you combine these.")
  } else {
    glue::glue("{L_name} and {S_name} are not the same length. To match the length of {L_name}, {S_name} will be repeated {times} times fully with the first {extra} elements repeated again. You will get a warning.")
  }
}

# Print test cases to show that your function fully works **in all cases**
recycling_checker(1:4, 1:2)  
recycling_checker(1:2, 1:4)   
recycling_checker(1:3, 1:3)  
recycling_checker(1:11, 1:3)  
recycling_checker(1:3, 1:11)   
```





## Exercise 3: `days_to_birthday()`

- **Function purpose:** Display the number of days from a user-supplied day to the user's next birthday. For example, if the user enters the day "Dec 24 2021" and a birthday of "January 1 2000," it should return a time difference of 8 days until the user's next birthday.  
- **Inputs:**
    - `birthday`: A string in month-day-year format giving the user's birthday
    - `from_when`: A string in month-day-year format. The number of days from this day to the user's upcoming birthday will be printed for the user. Set the default value for this argument to be `today()`.
- **Notes:**
    - Your function should be able to handle all of the following date formats: "Jan 1 2000", "January 1 2000", "Jan 1, 2000", "January 1, 2000", "1/1/2000", "1-1-2000". 
    - The two dates could be in different formats.
    - The `lubridate` [cheat sheet](https://rstudio.github.io/cheatsheets/lubridate.pdf) will be helpful.
    - You can subtract date objects with `-`.

```{r days_to_birthday}
# Write your function here
library(tidyverse)
library(lubridate)

days_to_birthday <- function(birthday, from_when = today()) {
  if (is.character(from_when)) {
    f_date <- parse_date_time(from_when, orders = c("mdy", "dmy", "ymd", "B d Y"))
  } else {
    f_date <- as_datetime(from_when)
  }

  b_date <- parse_date_time(birthday, orders = c("mdy", "B d Y", "b d Y", "m-d-y", "m/d/y"))
  
  if (is.na(b_date)) stop("Could not parse birthday format.")

  next_bday <- b_date
  year(next_bday) <- year(f_date)
  
  if (as.Date(next_bday) < as.Date(f_date)) {
    year(next_bday) <- year(f_date) + 1
  }
  
  diff <- as.numeric(difftime(next_bday, f_date, units = "days"))
  
    return(paste(ceiling(diff), "days until your next birthday!"))
}

# Include at least 6 test cases to show that your function fully works in all cases, as the Notes section describes.
days_to_birthday("Jan 1 2000", "Dec 24 2021")  
days_to_birthday("Jan 1 2000", "Jan 5 2021")   
days_to_birthday("October 31, 1995")          
days_to_birthday("12-25-1990", "2024-12-24")
```





## Exercise 4: `course_names_to_df()`

- **Function purpose:** Take a character vector of course numbers and names and store the different pieces of information in a data frame. See code chunk below for example output.
- **Inputs:**
    - `courses`: A character vector of course numbers and names. See code chunk below for an example.
    - `separate_cross_list`: A `TRUE/FALSE` indicator of whether a cross-listed course should have the two departments separated into 2 variables. Have the default value be `TRUE`.
- **Notes:**
    - Use regular expressions with the `str_extract()` function in the `stringr` package to write this function.

```{r course_names_to_df}
# Write your function here
library(tidyverse)

course_names_to_df <- function(courses, separate_cross_list = TRUE) {
  
  dept_raw    <- str_extract(courses, "^[^ ]+")
  course_num  <- as.numeric(str_extract(courses, "\\d+"))
  course_name <- str_extract(courses, "(?<=: ).*")
  
  results <- tibble(
    dept = dept_raw,
    course_num = course_num,
    course_name = course_name
  )
  
  if (separate_cross_list) {
    results <- results %>%
      separate(dept, into = c("dept1", "dept2"), sep = "/", fill = "right")
  }
  
  return(results)
}

# Include test cases to show that your function fully works
course_names <- c("STAT/COMP 212: Intermediate Data Science", "STAT 155: Introduction to Statistical Modeling", "LING 175: Sociolinguistics")

# The output for the above courses should look like the following:
# If separate_cross_list is TRUE
tibble(
    dept1 = c("STAT", "STAT", "LING"),
    dept2 = c("COMP", NA, NA),
    course_num = c(212, 155, 175),
    course_name = c("Intermediate Data Science", "Introduction to Statistical Modeling", "Sociolinguistics")
)
# If separate_cross_list is FALSE
tibble(
    dept = c("STAT/COMP", "STAT", "LING"),
    course_num = c(212, 155, 175),
    course_name = c("Intermediate Data Science", "Introduction to Statistical Modeling", "Sociolinguistics")
)


# Test your function with the test data above with separate_cross_list as TRUE and FALSE
course_names <- c("STAT/COMP 212: Intermediate Data Science", 
                  "STAT 155: Introduction to Statistical Modeling", 
                  "LING 175: Sociolinguistics")

course_names_to_df(course_names, separate_cross_list = TRUE)
course_names_to_df(course_names, separate_cross_list = FALSE)



```


## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- Lubridate parse_date_time reference
- Stringr str_extract documentation
- RStudio Regular Expressions cheatsheet


**Reflection:**

The lubridate documentation was critical for solving Exercise 3, as it helped me understand that parse_date_time requires a specific order of formats (like "B" for full month names) to handle the variety of user inputs. I found the logic for preventing negative birthday counts especially interesting, as it required using year() to shift dates into the future dynamically. Additionally, the Regex cheatsheet allowed me to use look-behind assertions (?<=) in Exercise 4, which made it much easier to extract course titles without including the leading colon from the raw text.
