---
title: "HW06 Functions + Iteration"
format: 
  html: 
    embed-resources: true
---

## Instructions

**Requirements:**

Ex 1

- Correct description for at least 3 of the \#1-\#4


Ex 2

- Correct description for at least 2 of the \#1-\#3


Ex 3

- Correctly modifies the code to work for all test cases provided


Ex 4

- Describes why the original code works
- Makes the function call clearer and more readable


Ex 5

- Correctly identifies the data type.
- Correctly describes all 3 c's roles in the code. 

Ex 6

- Correctly describe in comments the desired subsetting for at least 3.
- Correctly fixes the base R code to do the desired subsetting for at least 3.


Ex 7

- Correctly describe in words what the code is attempting to do and what it is doing (at a low level) for a numeric vector.
- Good guesses as to how it works with a tibble.


Ex 8

- Correctly extract elements from a fit model and the summary of a fit model. 



## Exercise 1

For each line marked with #1 to #4 in the code chunk below, describe in a sentence what the code is doing. 
```{r}
show_missing <- function(df, group_vars, summary_vars = everything()) {

    df |>
    group_by(pick({{ group_vars }})) |> #1
    summarize(
      across({{ summary_vars }}, ~ sum(is.na(.x))),
      .groups = "drop"
    ) |> #2
    select(where(~ any(.x > 0))) #3
}

nycflights13::flights |> show_missing(c(year, month, day)) #4
```

- \#1: The code groups the data frame by the variables specified in group_vars, using pick() to select multiple columns and {{ }} to unquote the argument passed by the user
- \#2: For each column selected by summary_vars, counts the number of NA values using across(), then drops the grouping structure with .groups = "drop"
- \#3: The code keeps only the columns where at least one value is greater than 0 meaning only columns that actually had missing values appear in the result
- \#4: Takes the flights dataset and calls show_missing() grouping by year, month, and day, so we can see which variables have missing values within each day


## Exercise 2

For each line marked with #1 to #3 in the code chunk below, describe in a sentence what the code is doing. 
```{r}
means <- c(1, 50, 20, 100, 120)

out <- vector("list", length(means)) #1
for (i in 1:length(means)) { #2
  out[[i]] <- rnorm(10, means[[i]]) #3
}
```

- \#1: Creates an empty list of the same length as means to store the results of each iteration
- \#2: Starts a for loop that runs once for each index position in means, from 1 to its length
- \#3: For each iteration i, generates 10 random normal values with mean means[[i]] and stores the result in the i-th slot of out


## Exercise 3

Consider the iteration in the code chunk below.  Modify it so that it works without any errors for all test cases listed at the top of the code chunk.
```{r}
means <- c(1, 50, 20, 100, 120)
# means <- c(1, 50, 20)
# means <- c()

out <- vector("list", length(means))
for (i in seq_along(means)) {
  out[[i]] <- rnorm(10, means[[i]])
}
```


## Exercise 4

Consider the code chunk below.  Describe in a sentence why code works then write a modified version below the original one that it is easier to read. 
```{r}
# Original
mean(, TRUE, x = c(1:10, NA))

# Modified
mean(x = c(1:10, NA), trim = 0, na.rm = TRUE)

```
- Why it works? 
R allows named arguments (x = c(1:10, NA)) to be placed out of position so the empty first slot is skipped and TRUE fills the next unmatched positional argument (trim), allowing the call to run without error

## Exercise 5

Consider the code chunk below:
```{r}
c <- 10
c(c = c)
```
What does the code return? Use language that describes the data type / class?

- It returns a named numeric vector of length 1, where the element has the name "c" and the value 10

Describe how each of the three `c`'s is interpreted and What is their role?

-  First c (left of <-): a variable name, it stores the value 10
- c(...) — the function c() which combines values into a vector
- c = c inside the function, the left c is the element name, a character label for the vector element,  the right c is the variable holding the value 10, which evaluates to 10


## Exercise 6

Each line in the code chunk below has a common data frame subsetting error.  First, add a comment above the line describing its intended subsetting purpose then fix the common subsetting errors.
```{r}
# Subset rows where number of cylinders equals 4
mtcars[mtcars$cyl == 4, ] 

# Subset rows 1 through 4 (exclude negative indexing mistake)
mtcars[1:4, ]

# Subset rows where cylinders are 5 or fewer
mtcars[mtcars$cyl <= 5, ]

# Subset rows where cylinders equal 4 or 6
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ]
```


## Exercise 7

In words, what does `df[is.na(df)] <- 0` attempt to do? 

- Replace all NA values in df with 0

How does it work if `df` is a numeric vector? Break the code down.

- is.na(df) returns a logical vector of the same length TRUE wherever df is NA, FALSE elsewhere
- df[is.na(df)] uses logical indexing to select only the NA positions
- <- 0 assigns 0 to those positions replacing the NAs in place

```{r}
df <- c(1:10,NA,11:20, NA)

```

Why does it work if `df` is a tibble with quantitative variables such as `mtcars`? 

- A data frame can be indexed like a matrix. is.na(df) produces a logical matrix of the same dimensions, df[logical_matrix] selects all cells where the condition is TRUE across all rows and columns and <- 0 replaces them all with 0. It works here because all columns are numeric so replacing with 0 is type-compatible

```{r}
df <- mtcars

#Adding missing values
df[c(1,3),c(2,4)] <- NA



```

Why do you get an error if `df <- nycflights13::flights`?

- The flights tibble contains factor columns (carrier, tailnum, origin, dest). When you try to assign 0 (a numeric value) into NA positions that belong to character columns R throws a type coercion error, we cannot replace a character NA with a numeric 0

```{r}
df <- nycflights13::flights

```


## Exercise 8

Given a linear model, e.g., `mod <- lm(mpg ~ wt, data = mtcars)`, extract the residual degrees of freedom. Then extract the R squared from the model summary (`summary(mod)`)
```{r}
mod <- lm(mpg ~ wt, data = mtcars)

mod$df.residual

summary(mod)$r.squared

```


## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- R documentation for seq_along(), mean(), and lm() function signatures
- dplyr pick() and across() documentation for dynamic column selection inside functions
- nycflights13 package documentation for understanding the flights dataset structure

**Reflection:**

The R documentation for seq_along() was essential for understanding why 1:length() fails on empty vectors, which was the core fix needed in Exercise 3. The mean() function signature documentation helped clarify how R matches named and positional arguments out of order, which made Exercise 4 much clearer to reason about. Understanding how across() and pick() work together inside a custom function was key for Exercise 1, as these tidyverse tools allow flexible column selection without hard-coding names. Finally, working through Exercise 7 reinforced how logical matrix indexing behaves differently across data types, which deepened my understanding of why the same operation succeeds on numeric vectors and tibbles but fails when character columns are present
