---
title: "Exam 1"
---

## Geodetic vs Projected Coordinate Reference Systems

A Coordinate Reference System (CRS) defines how spatial data are positioned on the Earth.

A geodetic CRS represents locations using latitude and longitude on the Earth's curved surface. Coordinates are expressed in degrees, which are angular units.

A projected CRS transforms the curved Earth into a flat coordinate system so that spatial calculations such as distance, area, and buffers can be computed using meters.

For example, WGS84 (EPSG:4326) is a geodetic CRS commonly used by GPS systems, while UTM projections are projected CRS commonly used for spatial analysis.


```{r}
library(sf)
library(tmap)
```

## Create a point in WGS84 (geodetic CRS)
```{r}
point <- st_sfc(
  st_point(c(-93.163, 44.953)), # St. Paul, MN
  crs = 4326
)

point

st_crs(point)
```

## Attempting a buffer in a geodetic CRS
```{r}
buffer_geo <- st_buffer(point, dist = 0.01)
```

## Transform to a projected CRS (UTM Zone 15N)
```{r}
point_proj <- st_transform(point, 26915)

st_crs(point_proj)
```

## Create a 1 km buffer
```{r}
buffer_proj <- st_buffer(point_proj, dist = 1000)
```

## Visualize the result
```{r}
tm_shape(buffer_proj) +
  tm_polygons() +
  tm_shape(point_proj) +
  tm_symbols(size = 0.5)
```

## Key takeaway

Geodetic CRS are useful for storing global location data, while projected CRS are necessary for accurate spatial measurements such as distance, area, and buffering.

Transforming between CRS using st_transform() is a standard step in spatial analysis workflows.


## Extra Exploration

## Compare distances
```{r}
point2 <- st_sfc(
  st_point(c(-93.10, 44.98)),
  crs = 4326
)

st_distance(point, point2)
```

```{r}
p1 <- st_transform(point, 26915)
p2 <- st_transform(point2, 26915)

st_distance(p1, p2)
```






