---
title: "HW03-P1 Adv Map Viz"
---


## Introduction

The following exercises will use census tract `tidycensus` data for Ramsey and Hennepin county and crash data (2019-2022) from the [pedestrian/bike crash database](https://information.stpaul.gov/datasets/stpaul::pedestrian-and-bike-crash-dataset-/about) for St. Paul within Ramsey county in the Twin Cities.

**Requirements:**

- Explore ACS and enter an addition variable to code
- Download and Read in roads data
- Ensure both census2020 and crashes are same CRS
- Crop roads (at least to Ramsey and Hennepin counties)
- Join crashes and census to calculate number of crashes per tract
- Join census2020 and crashes per tract
- Choropleth map of number of crashes per tract + use of scale_fill_gradientn()
- Choropleth map coloring census tracts by a census variable + crash locations as points + scale_fill_gradientn()
- Sentence Interpretation in Data Context
- 2 non-map visuals of number of crashes vs census variables
- Sentence Interpretation in Data Context


## 1 Load One More Variable

The following code loads census data for Ramsey counties and reads in the pedestrian/bike crash data.  Adjust this code to load in one additional census variable that you think might serve as useful context for crash data. To do this:

- Look at the documentation for `get_acs()` by entering `?tidycensus::get_acs` in the Console.
- Look at the [Seaching for variables documentation](https://walker-data.com/tidycensus/articles/basic-usage.html#searching-for-variables) from the `tidycensus` package. Use the function in this documentation to find one variable to add to the `variables` argument in the `get_acs()` function below.
- Rename the variable that you picked to have a more descriptive name within `rename()`.

> Enter the new variable you selected (e.g., B01003_001) here: B08301_019
> 
> Enter the new name for that variable (e.g., population) here: walk_commuters


```{r reading, message=FALSE, warning=FALSE, results="hide"}
library(tidyverse)
census2020 <- tidycensus::get_acs(
    year = 2020,
    state = "MN",
    geography = "tract",
    variables = c("B01003_001", "B19013_001", "B23006_023", "B25058_001", "B25107_001", "B25003_001", "B25003_002", "B25003_003", "B25077_001", "B08301_019"),
    output = "wide",
    geometry = TRUE
) |>
    filter(word(NAME, 4) %in% c("Ramsey")) |>
    mutate(
        tract = word(NAME, 3),
        tract = str_remove(tract, ","),
        county = word(NAME, 4)
    ) |>
    rename(
        "population" = "B01003_001E",
        "medianIncome" = "B19013_001E",
        "bachelors" = "B23006_023E",
        "medContractRent" = "B25058_001E",
        "tenureTotal" = "B25003_001E",
        "tenureOwned" = "B25003_002E",
        "tenureRented" = "B25003_003E",
        "medianHomeValue"= "B25077_001E",
        "walk_commuters" = "B08301_019E"
    ) |>
    dplyr::select(GEOID, population, medianIncome, bachelors, medContractRent, tenureTotal, tenureOwned, tenureRented, medianHomeValue, walk_commuters, tract, county, geometry)

crashes <- read_csv("https://mac-stat.github.io/data/Crash.csv") |>
    filter(!is.na(Latitude), !is.na(Longitude))
```

## 2 Download Shapefile

Visit [this site](https://catalog.data.gov/dataset/tiger-line-shapefile-2019-state-minnesota-primary-and-secondary-roads-state-based-shapefile) and download the Shapefile Zip File (`tl_2019_27_prisecroads.zip`) to the `data/raw` folder of this homework then unzip.

Load in the roads shapefile using `st_read()`, and transform `roads` to have the same CRS as `census2020` if necessary.

```{r reading_roads}
roads <- sf::st_read("data/raw/tl_2019_27_prisecroads/tl_2019_27_prisecroads.shp")

# Check CRS of roads and transform if necessary
roads <- sf::st_transform(roads, crs = sf::st_crs(census2020))
```


## 3 Transofrm to `sf` Object

Transform the `crashes` data frame to a `sf` object that has point geometry using the code below.

```{r crashes_to_sf}
library(sf)
crashes <- st_as_sf(crashes, coords = c("Longitude", "Latitude"), crs = "NAD83")
```


## 4 Check CRSs

Check if the CRSs are the same for `census2020` and `crashes` using `st_crs()`. Transform to ensure they are exactly the same.

```{r same_crs}
st_crs(census2020)
st_crs(crashes)

crashes <- st_transform(crashes, crs = st_crs(census2020))
st_crs(census2020) == st_crs(crashes)
```

## 5 Crop

Use `st_crop()` to crop the roads and census2020 datasets to the area we are interested in (crashes in St. Paul within Ramsey County). To determine the region to crop to, you will need to use `st_bbox()` on a spatial data object that only pertains to St. Paul within Ramsey County.

```{r crop_roads}
study_area_bbox <- st_bbox(crashes)
roads_sub <- st_crop(roads, study_area_bbox)
census2020_sub <- st_crop(census2020, study_area_bbox)
```

## 6 Join Spatial Datasets

Join the crashes and census dataset together and count the number of crashes per census tract. The function `st_join` can join to spatial data sets according to whether the spatial geometries of the right table intersect with the spatial geometries of the left table. It matters which order you join the datasets in!

```{r join_crashes}
crashes_per_tract <- st_join(crashes, census2020) |>
    st_drop_geometry() |> # removes geometry - makes the following calculation more efficient 
    filter(!is.na(Accident_Datetime)) |>
    count(GEOID, name = "n_crashes") 
```

## 7 Join Census

Join the census data with `crashes_per_tract` and then use a filter of `n > 0` to only keep the census tracts where crashes were recorded instead of all of Ramsey County. Again, order matters!

```{r add_geometry_back}
crashes_per_tract_geo <- census2020_sub |> # sf object with census geometry goes first
    left_join(crashes_per_tract, by = "GEOID") |>
    filter(n_crashes > 0)
```

## 8 Pedestrian/Bike Crashes per Census Tract

Create a map that shows the number of pedestrian/bike crashes per census tract. Write one sentence about the patterns you notices; interpret your findings in the data context.

- Use `scale_fill_gradientn()` to set up a meaningful color scale for the census tract colors.
- Include a map layer for the roads.

```{r map_num_crashes}
ggplot() +
  geom_sf(data = crashes_per_tract_geo, aes(fill = n_crashes), color = "white", size = 0.1) +
  geom_sf(data = roads_sub, color = "grey40", size = 0.2, alpha = 0.5) +
  
  scale_fill_gradientn(
    colors = c("yellow", "orange", "red", "darkred"),
    name = "Number of\nCrashes") +
  
  labs(
    title = "Pedestrian and Bike Crashes by Census Tract",
    subtitle = "St. Paul, MN (2019-2022)",
    caption = "Data: St. Paul Pedestrian/Bike Crash Dataset") +
  theme_minimal()
```

Interpretation: The map reveals a higher concentration of pedestrian and bike crashes in downtown St. Paul and along major transit corridors, suggesting that higher traffic density and pedestrian activity in these urban hubs correlate with increased accident frequency.

## 9 Crash Locations

Create one map that shows crash locations as points and colors census tracts by one different census variable. Write one sentence about the patterns you notice; interpret your findings in the data context.

- Use `scale_fill_gradientn()` to set up a color scale for the census tract colors.
- Include a layer for the roads.

```{r mas_crashes_census_info}
ggplot() +
  geom_sf(data = census2020_sub, aes(fill = walk_commuters), color = "white", size = 0.1) +
  geom_sf(data = roads_sub, color = "grey30", size = 0.2, alpha = 0.4) +
  geom_sf(data = crashes, color = "black", size = 0.5, alpha = 0.6) +
  
  scale_fill_gradientn(
    colors = c("lightblue", "blue", "darkblue"),
    name = "Commute by\nWalking") +
  
  labs(
    title = "St. Paul Crash Locations vs. Walking Populations",
    subtitle = "Points represent individual crashes; shading shows number of walking commuters",
    caption = "Data: ACS 2020 and St. Paul Open Data") +
  theme_minimal()
```

Interpretation: While crash points are distributed across the city, there is a visible clustering of incidents within census tracts that show a higher number of residents commuting by foot, particularly around the University Avenue corridor and downtown St. Paul.

## 10 Number of Crashes and Census Data

Create two other visuals (that are not maps) showing the relationships between the number of crashes and census data. Write one sentence about the patterns you notice; interpret your findings in the data context.

```{r crashes_census_info}
library(ggplot2)

# 1. Relationship between Walking Commuters and Number of Crashes
ggplot(crashes_per_tract_geo %>% st_drop_geometry(), 
       aes(x = walk_commuters, y = n_crashes)) +
  geom_point(alpha = 0.5, color = "darkblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +
  labs(title = "Crash Frequency vs. Walking Commuters",
       x = "Number of Walking Commuters",
       y = "Number of Crashes per Tract") +
  theme_minimal()

# 2. Relationship between Median Income and Number of Crashes
ggplot(crashes_per_tract_geo %>% st_drop_geometry(), 
       aes(x = medianIncome, y = n_crashes)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "loess", color = "orange", se = FALSE) +
  labs(title = "Crash Frequency vs. Median Income",
       x = "Median Household Income",
       y = "Number of Crashes per Tract") +
  theme_minimal()
```

Interpretation:

Visual 1 (Walking Commuters): There is a positive correlation between the number of walking commuters and crashes, suggesting that higher "pedestrian exposure" in a tract is a significant predictor of crash frequency.

Visual 2 (Median Income): The data shows a cluster of higher crash volumes in lower-to-middle income census tracts, potentially reflecting differences in infrastructure quality or transit dependency compared to wealthier areas.

## Resources Reflection (required)

List the resources you used to help with this assignment then write 3-5 sentence reflection on which resources were most helpful in finishing this task.

**Resources:**

- ggplot2 official documentation:
(https://ggplot2.tidyverse.org/reference)
- Patchwork package documentation: Used for organizing and combining multiple map and non-map visuals.
- RStudio ggplot2 cheatsheet: For looking up syntax for geom_sf and color scale functions.

**Reflection:**

The most helpful resource for this assignment was the combination of the ggplot2 documentation and real-time debugging for spatial joins. Learning how to properly sequence mutate(), rename(), and select() was a major breakthrough, as I realized that dropping the GEOID early in the process would break the entire spatial connection. I also found the sf package logic to be very powerful, seeing how crash points physically intersect with census polygons changed how I think about data "joining." Navigating the error messages taught me that spatial data requires much more attention to column management and Coordinate Reference Systems than standard data frames.
